%*******************************************************
% Abstract
%*******************************************************
%\renewcommand{\abstractname}{Abstract}
\pdfbookmark[1]{Abstract}{Abstract}
\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\chapter*{Abstract}
Functional RNA secondary structures are evolutionary conserved and can be predicted with various algorithms. Furthermore, RNA is able to fold onto itself during transcription, called the co-transcriptional folding event. The program kinwalker is able to predict such co-transcriptional foldings, by using a deterministic algorithm. The Kinwalker algorithm comes with different parameters, to fine tune said predictions. I performed a detailed parameter space analysis for 3 well studied RNA families. With the created data, evidence of evolutionary conservation of folding pathways can be investigated.




\vfill
\selectlanguage{ngerman}
\pdfbookmark[1]{Zusammenfassung}{Zusammenfassung}
\chapter*{Zusammenfassung}



Funktionelle RNA sekund{\"a}re Strukturen sind evolution{\"a}r konserviert, und k{\"o}nnen durch verschiedene Algorithmen vorherbestimmt werden. Dar{\"u}ber hinaus kann RNA w{\"a}rend der Transkription auf sich selbst zur{\"u}ckfalten. Dieses Ph{\"a}nomen wird auch Co-transkription{\"a}lles Falten genannt. Das Programm Kinwalker ist in der Lage, solch Co-transkription{\"a}lles Falten vorherzubestimmen. Kinwalker hat mehrere Parameter um Feinabstimmungen am Algorithmus zu t{\"a}tigen. Ich habe eine detailierte Parameter Raum Analyse get{\"a}tigt, mithilfe drei wissenschaftlich bekannter RNA Familien. Mit diesen erzeugten Daten, k{\"o}nnen Hinweise auf evolution{\"a}re Konservierung des Co-transkription{\"a}llen Faltungsvorgang untersucht werden.




\endgroup			

\vfill